Parsing Postfix log files

### 2007-09-23 doc/logparser.tex
Improved stats, added new graphs, improved explanations and updated source data. Added conclusions to parsing algorithm, parser coverage, limitations and graphs sections.

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.  XXX PROVIDE A
REFERENCE} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
IP address of the client is present in the listed RBL\@.

@@

The resulting execution times are plotted in graph~\ref{normal vs shuffled
vs reversed ordering}, where the difference is not readily apparent, but
graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
percentage increase of execution times:

\input{build/stats-normal-shuffle-reverse-percentage}

Overall this provides a modest but worthwhile performance increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\ref{normal regex vs discard regex} shows execution times with and
without caching the regex; also shown on the graph are the file sizes and
number of lines in each file.  Caching the compiled regexs is obviously far
more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
percentage execution time increase when not caching each regex.  Over all
93 log files the execution time increase when not caching each compiled
regex works out at:

XXX IMPROVE THIS\@.

\input{build/stats-cached-discarded-include-for-discussion}

@@

@@

@@

@@

Graphs are an excellent means of displaying data, transforming a
meaningless stream of numbers into an easily comprehensible form, where
anomalies and patterns are immediately obvious.  These graphs are used to
illustrate the topics discussed in section~\ref{rule efficiency}.  The
graphs in the first section cover parser scalability, demonstrating that
its performance scales linearly with input size.  The impact of rule
ordering is shown in section~\ref{rule ordering graphs}, and the lower
impact of rule ordering on larger log files is explained.  The final group
of graphs vividly shows the huge impact that caching compiled regexs has on
parser performance.

@@

\subsubsection{Execution time vs file size vs number of lines.}
\label{execution time vs file size vs number lines}
\includegraphics{build/plot-normal-filesize-numlines.ps}

\subsubsection{Ratio of file size and number of lines to execution time}
\label{execution time vs file size vs number lines factor}
\includegraphics{build/plot-normal-filesize-numlines-factor.ps}

\input{build/stats-normal-filesize-line-count-include}

@@

\subsubsection{Rule hits}
\label{rule hits}
\includegraphics{build/plot-hits.ps}

@@

\subsubsection{Log 22 distribution by rule}
\label{log22 distribution by rule}
\includegraphics{build/plot-log22-distribution-by-rule.ps}

@@

\subsubsection{Log 22 line and rule distribution by program}
\label{log22 line and rule distribution by program}
\includegraphics{build/plot-log22-distribution-versus-rule-count.ps}

@@

Two large dips can be seen in graph~\ref{normal regex vs disarded regex
factor} at log files 22 and 62--68, corresponding to the spikes in log file
size in graph~\ref{execution time vs file size vs number lines}.  The
anomaly at files 62--68 is caused by a combination of a mail loop and a
delivery failure which resulted in a huge number of log lines, all of which
were matched by a single rule.  Sorting the rules by the number of
successful matches has the effect that this rule will be tried first, so
the average number of regexs compiled per line drops sharply for those
eight log files, resulting in a much lower increase in relative execution
time, as detailed below.

@@

>>>>

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table \cite{cidr_table}}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

@@

The resulting execution times are plotted in graph~\ref{normal vs shuffled
vs reversed ordering}, where the difference is not readily apparent, but
graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
percentage increase of execution times (mean and standard deviation are
shown after the graph).  Overall this provides a modest but worthwhile
performance increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\ref{normal regex vs discard regex} shows execution times with and
without caching the regex.  Caching the compiled regexs is obviously far
more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
percentage execution time increase when not caching each regex.

Caching the compiled regexs is quite simple, and is the single most
effective optimisation tried in the parser.

@@

This section has presented the parsing algorithm, divided into: a high
level overview of the basic algorithm, followed by the first group of
difficulties thwarted.  The revised algorithm was detailed, preceded by a
flow chart depicting the paths a mail may take through the algorithm.  The
actions which are available to rules were described in full, and finally
the remaining complications discovered and overcome in the process of
refining and completing the algorithm.  Detecting, diagnosing and defeating
complications forms the largest single subsection of this document,
mirroring the development of the parser,  The complications are described
in the order they were overcome, with subsequent problems affecting fewer
mails (often by an order or magnitude), though the time required to solve
problems increased with each successive problem.

@@

Parser coverage is divided into two topics in this section: log lines
covered, and mails covered.  The former is initially more important, as the
parser must successfully parse every line if it is to be complete, but
subsequently the latter takes precedence because reproducing the path a
mail takes through Postfix is the aim of the parser.  Increasing the
percentage of log lines parsed is relatively simple and not intrusive:
adding new rules or modifying existing rules is simplified by the parser
implementation.  Improving the logical coverage is harder, as the actions
taken by Postfix must be reconstructed, and the new sequence of actions
integrated into the existing model without breaking the existing sequences.
detecting a deficiency in the parsing algorithm is also significantly
harder than detecting unparsed log lines, as the parser will warn about any
unparsed line, whereas the author must spot a flaw in the parser by
understanding the warnings generated.

@@

This section has covered the limitations of the parser and possible
improvements which may be implemented in the future.

@@

Graphs are an excellent means of displaying data, transforming a
meaningless stream of numbers into an easily comprehensible form, where
anomalies and patterns are immediately obvious.  These graphs are used to
illustrate the topics discussed in section~\ref{rule efficiency}.  The
graphs in the first section cover parser scalability, demonstrating that
performance scales linearly with input size.  The impact of rule ordering
is shown in section~\ref{rule ordering graphs}, and the lower impact of
rule ordering on larger log files is explained.  The final group of graphs
vividly shows the huge impact that caching compiled regexs has on parser
performance.

@@

\subsubsection{Execution time vs file size vs number of lines.}
\label{execution time vs file size vs number lines}

The Y axis represents the following:

    \item The time required, in seconds, to parse the log file.

    \item The size of the log file, in megabytes.

    \item The number of lines in the log file, divided by 10000.

\noindent
\includegraphics{build/plot-normal-filesize-numlines.ps}

\subsubsection{Ratio of file size and number of lines to execution time}
\label{execution time vs file size vs number lines factor}

\noindent
\includegraphics{build/plot-normal-filesize-numlines-factor.ps}

\input
\input{build/stats-normal-filesize-line-count-include}

@@

@@

\noindent\input{build/stats-normal-shuffle-reverse-include.tex}

\subsubsection{Log 22 distribution}
\label{log22 distribution by rule}
\includegraphics{build/plot-log22-distribution-by-rule.ps}

@@

\subsubsection{Log 22 line and rule distribution by program}
\label{log22 line and rule distribution by program}

\noindent
\includegraphics{build/plot-log22-distribution-versus-rule-count.ps}

\subsubsection{Log 22 lines per rule}
\label{log22 lines per rule}

As shown in this graph, the most common log lines (see graph~\ref{log22
line and rule distribution by program}) are from programs which have a low
number of rules, thus reducing the average parsing time per line, and the
overall execution time relative to the input size.

\noindent
\includegraphics{build/plot-log22-distribution-versus-rule-count-factor.ps}

\subsubsection{Rule hits}
\label{rule hits}

\noindent\includegraphics{build/plot-hits.ps}

@@

Two large dips can be seen in graph~\ref{normal regex vs disarded regex
factor} at log files 22 and 62--68, corresponding to the spikes in log file
size in graph~\ref{execution time vs file size vs number lines}.  The
anomalies are caused by a combination of a mail loop and a delivery failure
which resulted in a huge number of log lines, all of which were matched by
a single rule.  Sorting the rules by the number of successful matches has
the effect that this rule will be tried first, so the average number of
regexs compiled per line drops sharply for those eight log files, resulting
in a much lower increase in relative execution time, as detailed below.

@@

The graphs presented in this section illustrate the topics discussed in
section~\ref{rule efficiency}.  The first collection of graphs are about
parser scalability, showing the linearly relationship between execution
time and input size.  Section~\ref{rule ordering graphs} relates to the
effect of rule ordering on execution time, and the unexpected consequences
of specific inputs.  The necessity of caching compiled regexs is attested
to by the final group of graphs, where the difference between caching and
discarding is staggering.
